---
title: "Getting Started with gdpins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gdpins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # Drive operations need auth; keep chunks non-executable in CI
)
```

`gdpins` is a package that layers **Google Drive + pins** for reproducible,
offline-capable data management. Drive is the source of truth; a local cache
mirrors it. Reads are local-first; writes fan out; sync and delete are
explicit and guarded.

## Installation

`gdpins` is a package-in-place — load it from the project root:

```{r load}
pkgload::load_all(".")
# or, once installed:
# library(gdpins)
```

## Authentication

Drive operations use lazy, CLI-guided auth. Set your email once in
`.Renviron`:

```
GDRIVE_EMAIL=you@example.com
```

Check connectivity:

```{r online}
gdpins_is_online()
```

## Drive adapter

All Drive operations go through an adapter. Create one with `gdpins_real_drive()`
by supplying your **project root folder ID** (visible in the folder's Drive URL):

```
https://drive.google.com/drive/folders/<your-folder-id-here>
```

```{r adapter}
# Picks up GDRIVE_EMAIL by default:
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")

# Or a path string (resolved automatically via googledrive::drive_get):
adapter <- gdpins_real_drive("My Drive/kazLandEconImpact-data")

# Optional explicit override:
adapter <- gdpins_real_drive(
  "1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms",
  email = "you@example.com"
)
```

Optional pre-flight auth call (usually unnecessary if you construct a real
adapter right away):

```{r auth}
gdpins_ensure_drive_auth()
```

### Verify folder locations

Before wiring up boards, confirm you have the right folder:

```{r drive-url}
# Returns a clickable browser URL for any folder in the adapter root
gdpins_drive_url(adapter)                              # root folder
gdpins_drive_url(adapter, "kazLandEconImpact-data/data-raw")  # subfolder
```

You can also pass a Drive folder ID directly as `drive_path` when you know
exactly which folder to use — `gdpins` will verify it exists and error
immediately if not:

```{r adapter-id-as-path}
bd_raw <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "1AbCdEfGhIjKlMnOpQrStUvWxYz0123456",  # folder ID directly
  cache_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"), "data-raw-cache"),
  adapter    = adapter
)
```

## Board initialisation

A board wraps a layer of the data pipeline. Three configurations:

```{r board-local-only}
# Local only — no Drive, works fully offline (no adapter needed)
bd_local <- gdpins_init_board(
  name      = "scratch",
  local_dir = file.path(Sys.getenv("OD_PRIV_ROOT"), "scratch")
)
```

```{r board-drive-cache}
# Drive + cache — standard production configuration
bd_raw <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "kazLandEconImpact-data/data-raw",
  cache_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "data-raw-cache"),
  adapter    = adapter,
  versioned  = TRUE,
  create     = NA,          # prompt to create if absent
  on_discrepancy = NULL     # "prompt" interactively / "warn" in scripts
)
```

```{r board-super}
# Drive + cache + standalone local (super config)
bd_clean <- gdpins_init_board(
  name       = "data_clean",
  drive_path = "kazLandEconImpact-data/data-clean",
  cache_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "data-clean-cache"),
  local_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "data-clean-local"),
  adapter    = adapter,
  versioned  = TRUE
)
```

## Writing and reading pins

Pin names are **bare snake_case** — the board encodes the layer.

```{r pin-write}
library(dplyr)

# Any R object — tibble, sf, list, nested tibble
gdpins_pin_write(bd_raw, my_tibble, "gdp_panel")
gdpins_pin_write(bd_raw, my_sf,     "parcels")    # sf auto-encoded as parquet
```

```{r pin-read}
gdp_panel <- gdpins_pin_read(bd_raw, "gdp_panel")
parcels   <- gdpins_pin_read(bd_raw, "parcels")   # sf decoded with CRS intact
```

```{r pin-remove}
# Remove a pin across all configured board layers (idempotent if missing)
gdpins_pin_remove(bd_raw, "gdp_panel")
```

## Format selection

`gdpins` auto-detects the storage format via `gdpins_detect_format()`. The
rules, in order:

| Object type | Chosen format | Storage |
|-------------|--------------|---------|
| Not a data frame (list, model, …) | `"rds"` | `saveRDS()` |
| Data frame with list-columns (non-sf) | `"rds"` | `saveRDS()` |
| Plain tibble / data frame | `"arrow"` | Parquet via `arrow` |
| `sf` object (all list-cols are `sfc`) | `"arrow"` | Parquet (geometry WKT-encoded) |

```{r format-auto}
gdpins_detect_format(mtcars)         # "arrow"  — plain data frame
gdpins_detect_format(my_sf)          # "arrow"  — sf, sfc cols are fine
gdpins_detect_format(my_model)       # "rds"    — not a data frame
gdpins_detect_format(nested_tibble)  # "rds"    — has non-sfc list-column
```

Override the format explicitly when you need to:

```{r format-explicit}
# Force RDS — useful for tibbles with custom attributes you want to preserve
gdpins_pin_write(bd_raw, my_tibble, "gdp_panel", format = "rds")

# Force Arrow/Parquet — rarely needed; auto-detect handles most cases
gdpins_pin_write(bd_raw, my_tibble, "gdp_panel", format = "arrow")
```

> **Rule of thumb:** let auto-detection choose. Override to `"rds"` only if
> the object has list-columns with non-spatial content (e.g. nested model
> output) that you want to read back in R without re-building.

## Spatial data (sf) round-trip

`sf` objects are stored as **Parquet** with geometry converted to WKT text.
CRS (EPSG code) is embedded in the column name using the suffix `__<epsg>__`:

```
geometry column "geometry" with EPSG 4326
  → stored as column "geometry__4326__" (WKT text)
  → restored as column "geometry" (sfc) with CRS 4326 on read
```

All of this happens automatically — no manual encoding/decoding needed:

```{r sf-write-read}
# Write sf — geometry and CRS encoded automatically
gdpins_pin_write(bd_raw, parcels_sf, "parcels")

# Read back — sf class and CRS fully restored
parcels <- gdpins_pin_read(bd_raw, "parcels")
sf::st_crs(parcels)$epsg  # original EPSG

# Multiple geometry columns with different CRS are each encoded separately
gdpins_pin_write(bd_raw, multi_geom_sf, "parcels_multi")
```

The helpers `gdpins_sf_to_parquet()` and `gdpins_parquet_to_sf()` are exposed
if you need to encode/decode manually (e.g. to write to an external Parquet file):

```{r sf-manual}
encoded <- gdpins_sf_to_parquet(parcels_sf)   # tibble with geometry__4326__ col
restored <- gdpins_parquet_to_sf(encoded)      # sf with CRS restored
```

> **Limitation:** geometry columns must have an EPSG integer CRS. Custom
> proj-string-only CRS will error. Set a CRS first with `sf::st_set_crs()`.

## Reading a specific version

When `versioned = TRUE` (the default), each `gdpins_pin_write()` call creates
a new version rather than overwriting. Versions are labelled by timestamp.

```{r versions-list}
# Inspect all versions of a pin
info <- gdpins_pin_info(bd_raw, "gdp_panel")
info$versions        # tibble: version label, created, size
info$n_versions      # integer count
```

```{r versions-read}
# Read the latest version (default)
gdpins_pin_read(bd_raw, "gdp_panel")

# Read a specific historical version by its label
gdpins_pin_read(bd_raw, "gdp_panel", version = "20240615T120000Z-abc12")
```

```{r versions-prune}
# Keep only the 3 most recent versions, dry-run first
gdpins_prune_pin_versions(bd_raw, "gdp_panel", keep = 3, dry_run = TRUE)
gdpins_prune_pin_versions(bd_raw, "gdp_panel", keep = 3, dry_run = FALSE)
```

> For a `versioned = FALSE` board, each write **overwrites** the previous
> value. Use this for frequently-updated reference data where history is
> unwanted.

## Offline behaviour

When Drive is unreachable (no internet, VPN down), `gdpins_init_board()` falls
back automatically:

| Config | Offline behaviour |
|--------|------------------|
| `local_only` | Fully offline — no change |
| `drive_cache` | Falls back to cache directory as local-only board |
| `drive_cache_local` | Falls back to standalone `local_dir` board |

A warning is emitted; writes are blocked (Drive boards are write-protected
offline); reads continue from the local copy.

```{r offline}
# Check connectivity explicitly before Drive operations
gdpins_is_online()   # TRUE / FALSE

# Board status shows sync state between Drive and local
gdpins_board_status(bd_raw)

# Reads work offline — served from cache or local_dir
gdp_panel <- gdpins_pin_read(bd_raw, "gdp_panel")  # no network needed
```

The fallback above is automatic and happens only when Drive is unreachable at
init time. To *deliberately* detach an already-connected board or raw
connection from Drive — e.g. to avoid network calls for a while, or to keep
working through a flaky connection — use `gdpins_go_offline()` and
`gdpins_go_online()` instead. See "Deliberately going offline and back
online" in `vignette("google-drive-adapter")` for details.

## Raw-exogenous connection

For data as received from APIs — plain files, no pins metadata.

```{r raw-connect}
conn <- gdpins_raw_connect(
  drive_path = "kazLandEconImpact-data/raw-exogenous/worldbank-api",
  local_path = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "raw-exogenous",
                         "worldbank-api"),
  adapter    = adapter,
  create     = TRUE
)
```

```{r raw-verbs}
# Store an R object (serialised to the appropriate format by extension)
gdpins_raw_put_object(conn, api_response_tbl, "gdp_2024.parquet")

# Upload a file verbatim (byte-faithful, any extension)
gdpins_raw_put_file(conn, path = "downloads/kadaster_raw.geojson",
                    name = "kadaster_raw.geojson")

# Retrieve (reads from local mirror; Drive only if absent locally)
gdp_raw  <- gdpins_raw_get(conn, "gdp_2024.parquet")
kad_raw  <- gdpins_raw_get(conn, "kadaster_raw.geojson")

# List contents — 8-column tibble including local paths and Drive IDs
listing <- gdpins_raw_ls(conn, depth = 2)
# listing$local_path  — absolute path on this machine
# listing$drive_id    — Drive file/folder ID (NA on fake adapter)
# listing$drive_url   — clickable browser URL (NA on fake adapter)

# Resolve a file to its local path, downloading from Drive if needed
local_path <- gdpins_raw_path(conn, "gdp_2024.parquet")
arrow::read_parquet(local_path)  # pass to any reader

# Non-standard filenames (spaces, parentheses) work too
gdpins_raw_path(conn, "quarterly report (Q1 2024).csv")

# Remove one file (flat or nested path). Missing files are ignored.
gdpins_raw_remove(conn, "kadaster_raw.geojson")
```

## Sync

Sync is **always explicit** — never automatic.

```{r sync}
# Auto: newer side wins, both directions
gdpins_sync(bd_raw)

# Explicit direction
gdpins_sync(bd_raw, direction = "from_drive")   # pull Drive → local
gdpins_sync(bd_raw, direction = "to_drive")     # push local → Drive

# Check status first
gdpins_board_status(bd_raw)
```

Conflict handling for versioned boards: both sides simply become new versions
(zero data loss). For raw/unversioned boards: `on_conflict = "stop"` halts
and reports; `"prompt"` asks per conflict.

## Discovery

```{r discovery}
# List all pins
gdpins_list_pins(bd_clean)

# Detailed info for one pin
gdpins_pin_info(bd_clean, "parcels")
```

## Output

```{r output}
library(ggplot2)

# Save a figure (PNG or SVG — ggplot object is NOT stored)
gdpins_save_figure(
  plot   = ggplot(gdp_panel, aes(year, value)) + geom_line(),
  name   = "gdp_trend",
  dir    = here::here("output", "figures"),
  width  = 8,
  height = 5,
  device = "png"
)

# Write output table to output board
gdpins_pin_write(bd_output, summary_tbl, "regional_summary")

# Publish local output to Drive (deliberate final step)
gdpins_publish_output(
  tables_board  = bd_output,
  figures_dir   = here::here("output", "figures"),
  drive_tables  = "output-tables",
  drive_figures = "output-figures",
  adapter       = adapter,
  dry_run       = TRUE   # preview first
)

# Then actually publish:
gdpins_publish_output(
  tables_board  = bd_output,
  figures_dir   = here::here("output", "figures"),
  adapter       = adapter,
  dry_run       = FALSE
)
```

## Pruning old versions

```{r prune}
# Preview what would be removed (dry_run = TRUE by default)
gdpins_prune_pin_versions(bd_raw, "gdp_panel", keep = 3, dry_run = TRUE)

# Actually prune
gdpins_prune_board_versions(bd_raw, keep = 2, dry_run = FALSE, force = TRUE)
```

## Console output

`gdpins` objects print compactly (≤80 cols):

```{r print}
print(bd_raw)
summary(bd_raw)
gdpins_list_pins(bd_raw)
gdpins_pin_info(bd_raw, "gdp_panel")
```

## Layered data flow

```
API/source  ─►  raw-exogenous  ─►  data-raw  ─►  data-interm  ─►  data-clean
  (gdpins_raw_*)                    (gdpins_pin_write / gdpins_pin_read on each board)
                                                                        ▼
                                          output-tables board + output-figures folder
                                                   (gdpins_publish_output)
```

Pin names are bare snake_case. The board encodes the layer. A concept keeps
its name (`parcels`, `gdp_panel`) as it flows `raw → interm → clean`,
giving readable lineage without renaming.

See [`docs/adr/0001-gdpins-data-architecture.md`](../docs/adr/0001-gdpins-data-architecture.md)
for the full architectural decision record.

See also the **[Google Drive Adapter](google-drive-adapter.html)** vignette for
a deeper guide to adapter setup, folder ID lookup, and `gdpins_drive_url()`.
