---
title: "Getting Started with gdpins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gdpins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # Drive operations need auth; keep chunks non-executable in CI
)
```

`gdpins` is a package that layers **Google Drive + pins** for reproducible,
offline-capable data management. Drive is the source of truth; a local cache
mirrors it. Reads are local-first; writes fan out; sync and delete are
explicit and guarded.

## Installation

`gdpins` is a package-in-place — load it from the project root:

```{r load}
pkgload::load_all(".")
# or, once installed:
# library(gdpins)
```

## Authentication

Drive operations use lazy, CLI-guided auth. Set your email once in
`.Renviron`:

```
GDRIVE_EMAIL=you@example.com
```

Then auth is triggered automatically on the first Drive operation, or call it
explicitly:

```{r auth}
gdpins_ensure_drive_auth()
```

Check connectivity:

```{r online}
gdpins_is_online()
```

## Drive adapter

All Drive operations go through an adapter. Create one with `gdpins_real_drive()`
by supplying your **project root folder ID** (visible in the folder's Drive URL):

```
https://drive.google.com/drive/folders/<your-folder-id-here>
```

```{r adapter}
# Accepts a Drive folder ID:
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")

# Or a path string (resolved automatically via googledrive::drive_get):
adapter <- gdpins_real_drive("My Drive/kazLandEconImpact-data")
```

### Verify folder locations

Before wiring up boards, confirm you have the right folder:

```{r drive-url}
# Returns a clickable browser URL for any folder in the adapter root
gdpins_drive_url(adapter)                              # root folder
gdpins_drive_url(adapter, "kazLandEconImpact-data/data-raw")  # subfolder
```

You can also pass a Drive folder ID directly as `drive_path` when you know
exactly which folder to use — `gdpins` will verify it exists and error
immediately if not:

```{r adapter-id-as-path}
bd_raw <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "1AbCdEfGhIjKlMnOpQrStUvWxYz0123456",  # folder ID directly
  cache_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"), "data-raw-cache"),
  adapter    = adapter
)
```

## Board initialisation

A board wraps a layer of the data pipeline. Three configurations:

```{r board-local-only}
# Local only — no Drive, works fully offline (no adapter needed)
bd_local <- gdpins_init_board(
  name      = "scratch",
  local_dir = file.path(Sys.getenv("OD_PRIV_ROOT"), "scratch")
)
```

```{r board-drive-cache}
# Drive + cache — standard production configuration
bd_raw <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "kazLandEconImpact-data/data-raw",
  cache_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "data-raw-cache"),
  adapter    = adapter,
  versioned  = TRUE,
  create     = NA,          # prompt to create if absent
  on_discrepancy = NULL     # "prompt" interactively / "warn" in scripts
)
```

```{r board-super}
# Drive + cache + standalone local (super config)
bd_clean <- gdpins_init_board(
  name       = "data_clean",
  drive_path = "kazLandEconImpact-data/data-clean",
  cache_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "data-clean-cache"),
  local_dir  = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "data-clean-local"),
  adapter    = adapter,
  versioned  = TRUE
)
```

## Writing and reading pins

Pin names are **bare snake_case** — the board encodes the layer.

```{r pin-write}
library(dplyr)

# Any R object — tibble, sf, list, nested tibble
gdpins_pin_write(bd_raw, my_tibble, "gdp_panel")
gdpins_pin_write(bd_raw, my_sf,     "parcels")    # sf auto-encoded as parquet
```

```{r pin-read}
gdp_panel <- gdpins_pin_read(bd_raw, "gdp_panel")
parcels   <- gdpins_pin_read(bd_raw, "parcels")   # sf decoded with CRS intact
```

```{r pin-remove}
# Remove a pin across all configured board layers (idempotent if missing)
gdpins_pin_remove(bd_raw, "gdp_panel")
```

## Raw-exogenous connection

For data as received from APIs — plain files, no pins metadata.

```{r raw-connect}
conn <- gdpins_raw_connect(
  drive_path = "kazLandEconImpact-data/raw-exogenous/worldbank-api",
  local_path = file.path(Sys.getenv("OD_PRIV_ROOT"),
                         "kazLandEconImpact-data", "raw-exogenous",
                         "worldbank-api"),
  adapter    = adapter,
  create     = TRUE
)
```

```{r raw-verbs}
# Store an R object (serialised to the appropriate format by extension)
gdpins_raw_put_object(conn, api_response_tbl, "gdp_2024.parquet")

# Upload a file verbatim (byte-faithful)
gdpins_raw_put_file(conn, path = "downloads/kadaster_raw.geojson",
                    name = "kadaster_raw.geojson")

# Retrieve (reads from local mirror; Drive only if absent locally)
gdp_raw  <- gdpins_raw_get(conn, "gdp_2024.parquet")
kad_raw  <- gdpins_raw_get(conn, "kadaster_raw.geojson")

# List contents
gdpins_raw_ls(conn, depth = 2)

# Remove one file (flat or nested path). Missing files are ignored.
gdpins_raw_remove(conn, "kadaster_raw.geojson")
```

## Sync

Sync is **always explicit** — never automatic.

```{r sync}
# Auto: newer side wins, both directions
gdpins_sync(bd_raw)

# Explicit direction
gdpins_sync(bd_raw, direction = "from_drive")   # pull Drive → local
gdpins_sync(bd_raw, direction = "to_drive")     # push local → Drive

# Check status first
gdpins_board_status(bd_raw)
```

Conflict handling for versioned boards: both sides simply become new versions
(zero data loss). For raw/unversioned boards: `on_conflict = "stop"` halts
and reports; `"prompt"` asks per conflict.

## Discovery

```{r discovery}
# List all pins
gdpins_list_pins(bd_clean)

# Detailed info for one pin
gdpins_pin_info(bd_clean, "parcels")
```

## Output

```{r output}
library(ggplot2)

# Save a figure (PNG or SVG — ggplot object is NOT stored)
gdpins_save_figure(
  plot   = ggplot(gdp_panel, aes(year, value)) + geom_line(),
  name   = "gdp_trend",
  dir    = here::here("output", "figures"),
  width  = 8,
  height = 5,
  device = "png"
)

# Write output table to output board
gdpins_pin_write(bd_output, summary_tbl, "regional_summary")

# Publish local output to Drive (deliberate final step)
gdpins_publish_output(
  tables_board  = bd_output,
  figures_dir   = here::here("output", "figures"),
  drive_tables  = "output-tables",
  drive_figures = "output-figures",
  adapter       = adapter,
  dry_run       = TRUE   # preview first
)

# Then actually publish:
gdpins_publish_output(
  tables_board  = bd_output,
  figures_dir   = here::here("output", "figures"),
  adapter       = adapter,
  dry_run       = FALSE
)
```

## Pruning old versions

```{r prune}
# Preview what would be removed (dry_run = TRUE by default)
gdpins_prune_pin_versions(bd_raw, "gdp_panel", keep = 3, dry_run = TRUE)

# Actually prune
gdpins_prune_board_versions(bd_raw, keep = 2, dry_run = FALSE, force = TRUE)
```

## Console output

`gdpins` objects print compactly (≤80 cols):

```{r print}
print(bd_raw)
summary(bd_raw)
gdpins_list_pins(bd_raw)
gdpins_pin_info(bd_raw, "gdp_panel")
```

## Layered data flow

```
API/source  ─►  raw-exogenous  ─►  data-raw  ─►  data-interm  ─►  data-clean
  (gdpins_raw_*)                    (gdpins_pin_write / gdpins_pin_read on each board)
                                                                        ▼
                                          output-tables board + output-figures folder
                                                   (gdpins_publish_output)
```

Pin names are bare snake_case. The board encodes the layer. A concept keeps
its name (`parcels`, `gdp_panel`) as it flows `raw → interm → clean`,
giving readable lineage without renaming.

See [`docs/adr/0001-gdpins-data-architecture.md`](../docs/adr/0001-gdpins-data-architecture.md)
for the full architectural decision record.

See also the **[Google Drive Adapter](google-drive-adapter.html)** vignette for
a deeper guide to adapter setup, folder ID lookup, and `gdpins_drive_url()`.
