---
title: "Setting up the Google Drive Adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up the Google Drive Adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gdpins)
```

## What is the Drive adapter?

The `gdpins` package isolates all Google Drive operations behind a *Drive adapter* (`gdpins_drive_adapter`). Every function that touches Drive — boards, raw connections, sync, publish — receives the adapter rather than calling `googledrive` directly. This seam lets you swap a real adapter for a fake (tempdir-backed) one in tests without any network calls.

## Creating a real adapter

Use `gdpins_real_drive()` to create a real adapter. You must supply the **Drive folder ID** of your project root folder — or a path string to look it up.

### Finding your folder ID

Open the folder in Google Drive in a browser. The URL looks like:

```
https://drive.google.com/drive/folders/1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms
```

The last segment is the folder ID.

### Creating the adapter

```{r, eval=FALSE}
# Using a Drive folder ID (recommended):
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")

# Using a path string (resolved via googledrive::drive_get):
adapter <- gdpins_real_drive("My Drive/projects/myproject")
```

## Verifying the folder location

Use `gdpins_drive_url()` to get a clickable URL for any folder within the adapter root:

```{r, eval=FALSE}
# Root folder URL:
gdpins_drive_url(adapter)

# Subfolder URL:
gdpins_drive_url(adapter, "data-raw")
```

Copy-paste the URL into your browser to visually confirm the location before connecting boards.

## Setting up a board

```{r, eval=FALSE}
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "my-project/data-raw",   # relative to adapter root
  cache_dir  = "~/.cache/gdpins/data-raw",
  adapter    = adapter,
  create     = TRUE
)
board
```

You can also supply a Drive folder ID directly as `drive_path` — useful when you know the exact folder ID:

```{r, eval=FALSE}
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "1AbCdEfGhIjKlMnOpQrStUvWxYz0123456",
  cache_dir  = "~/.cache/gdpins/data-raw",
  adapter    = adapter
)
```

When `drive_path` is a folder ID, `gdpins` verifies it exists and errors immediately if not found. The `create` argument is ignored — folder IDs cannot be auto-created.

## Setting up a raw connection

```{r, eval=FALSE}
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = "data/raw/worldbank-api",
  adapter    = adapter,
  create     = TRUE
)
conn
```

## Testing with a fake adapter

In scripts and packages, use `gdpins_fake_drive()` for offline testing:

```{r}
adapter <- gdpins_fake_drive()

board <- gdpins_init_board(
  name       = "test_board",
  drive_path = "project/data-raw",
  cache_dir  = tempfile("cache_"),
  adapter    = adapter,
  create     = TRUE
)
board
```

The fake adapter mirrors Drive operations on the local filesystem — no credentials, no network, fully reproducible.
