---
title: "Setting up the Google Drive Adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up the Google Drive Adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gdpins)
```

## What is the Drive adapter?

The `gdpins` package isolates all Google Drive operations behind a *Drive adapter* (`gdpins_drive_adapter`). Every function that touches Drive — boards, raw connections, sync, publish — receives the adapter rather than calling `googledrive` directly. This seam lets you swap a real adapter for a fake (tempdir-backed) one in tests without any network calls.

## Creating a real adapter

Use `gdpins_real_drive()` to create a real adapter. You must supply the **Drive folder ID** of your project root folder — or a path string to look it up.

### Finding your folder ID

Open the folder in Google Drive in a browser. The URL looks like:

```
https://drive.google.com/drive/folders/1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms
```

The last segment is the folder ID.

### Creating the adapter

```{r, eval=FALSE}
# Using a Drive folder ID (recommended):
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")

# Using a path string (resolved via googledrive::drive_get):
adapter <- gdpins_real_drive("My Drive/projects/myproject")
```

### Authentication and the `email` parameter

`gdpins_real_drive()` now calls `gdpins_ensure_drive_auth()` automatically.

```{r, eval=FALSE}
# 1) Uses GDRIVE_EMAIL from .Renviron (recommended)
adapter <- gdpins_real_drive("folder-id")

# 2) Explicit override
adapter <- gdpins_real_drive("folder-id", email = "explicit@example.com")

# 3) No email set -> gargle interactive CLI account picker
adapter <- gdpins_real_drive("folder-id", email = "")
```

Set this once in `~/.Renviron`:

```txt
GDRIVE_EMAIL=you@example.com
```

An explicit pre-flight `gdpins_ensure_drive_auth()` call is still valid, but
optional when creating adapters with `gdpins_real_drive()`.

## Verifying the folder location

Use `gdpins_drive_url()` to get a clickable URL for any folder within the adapter root:

```{r, eval=FALSE}
# Root folder URL:
gdpins_drive_url(adapter)

# Subfolder URL:
gdpins_drive_url(adapter, "data-raw")
```

Copy-paste the URL into your browser to visually confirm the location before connecting boards.

## Setting up a board

```{r, eval=FALSE}
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "my-project/data-raw",   # relative to adapter root
  cache_dir  = "~/.cache/gdpins/data-raw",
  adapter    = adapter,
  create     = TRUE
)
board
```

You can also supply a Drive folder ID directly as `drive_path` — useful when you know the exact folder ID:

```{r, eval=FALSE}
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "1AbCdEfGhIjKlMnOpQrStUvWxYz0123456",
  cache_dir  = "~/.cache/gdpins/data-raw",
  adapter    = adapter
)
```

When `drive_path` is a folder ID, `gdpins` verifies it exists and errors immediately if not found. The `create` argument is ignored — folder IDs cannot be auto-created.

## Setting up a raw connection

```{r, eval=FALSE}
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = "data/raw/worldbank-api",
  adapter    = adapter,
  create     = TRUE
)
conn
```

Nested uploads via `gdpins_raw_put_object()` and `gdpins_raw_put_file()` create
missing intermediate Drive folders automatically (for example
`"sub/sub/folder/file.rds"`).

## Removal behavior (full vs partial)

`gdpins` supports explicit, idempotent remove verbs:

```{r, eval=FALSE}
# Boards: removes a pin from all configured board layers
gdpins_pin_remove(board, "cars")

# Raw connections: removes one file (flat or nested path)
gdpins_raw_remove(conn, "sub/sub/file.rds")
```

Drive removals are recoverable: files are moved to Drive trash (not hard-deleted).
For raw connections, removal scope is single files only; recursive folder deletion
is intentionally unsupported.

## Testing with a fake adapter

In scripts and packages, use `gdpins_fake_drive()` for offline testing:

```{r}
adapter <- gdpins_fake_drive()

board <- gdpins_init_board(
  name       = "test_board",
  drive_path = "project/data-raw",
  cache_dir  = tempfile("cache_"),
  adapter    = adapter,
  create     = TRUE
)
board
```

The fake adapter mirrors Drive operations on the local filesystem — no credentials, no network, fully reproducible.
