---
title: "Setting up the Google Drive Adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up the Google Drive Adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gdpins)
```

## What is the Drive adapter?

The `gdpins` package isolates all Google Drive operations behind a *Drive adapter* (`gdpins_drive_adapter`). Every function that touches Drive — boards, raw connections, sync, publish — receives the adapter rather than calling `googledrive` directly. This seam lets you swap a real adapter for a fake (tempdir-backed) one in tests without any network calls.

## Creating a real adapter

Use `gdpins_real_drive()` to create a real adapter. You must supply the **Drive folder ID** of your project root folder — or a path string to look it up.

### Finding your folder ID

Open the folder in Google Drive in a browser. The URL looks like:

```
https://drive.google.com/drive/folders/1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms
```

The last segment is the folder ID.

### Creating the adapter

```{r, eval=FALSE}
# Using a Drive folder ID (recommended):
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")

# Using a path string (resolved via googledrive::drive_get):
adapter <- gdpins_real_drive("My Drive/projects/myproject")
```

### Authentication and the `email` parameter

`gdpins_real_drive()` now calls `gdpins_ensure_drive_auth()` automatically.

```{r, eval=FALSE}
# 1) Uses GDRIVE_EMAIL from .Renviron (recommended)
adapter <- gdpins_real_drive("folder-id")

# 2) Explicit override
adapter <- gdpins_real_drive("folder-id", email = "explicit@example.com")

# 3) No email set -> gargle interactive CLI account picker
adapter <- gdpins_real_drive("folder-id", email = "")
```

Set this once in `~/.Renviron`:

```txt
GDRIVE_EMAIL=you@example.com
```

An explicit pre-flight `gdpins_ensure_drive_auth()` call is still valid, but
optional when creating adapters with `gdpins_real_drive()`.

## Verifying the folder location

Use `gdpins_drive_url()` to get a clickable URL for any folder within the adapter root:

```{r, eval=FALSE}
# Root folder URL:
gdpins_drive_url(adapter)

# Subfolder URL:
gdpins_drive_url(adapter, "data-raw")
```

Copy-paste the URL into your browser to visually confirm the location before connecting boards.

## Setting up a board

```{r, eval=FALSE}
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "my-project/data-raw",   # relative to adapter root
  cache_dir  = "~/.cache/gdpins/data-raw",
  adapter    = adapter,
  create     = TRUE
)
board
```

You can also supply a Drive folder ID directly as `drive_path` — useful when you know the exact folder ID:

```{r, eval=FALSE}
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "1AbCdEfGhIjKlMnOpQrStUvWxYz0123456",
  cache_dir  = "~/.cache/gdpins/data-raw",
  adapter    = adapter
)
```

When `drive_path` is a folder ID, `gdpins` verifies it exists and errors immediately if not found. The `create` argument is ignored — folder IDs cannot be auto-created.

## Setting up a raw connection

```{r, eval=FALSE}
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = "data/raw/worldbank-api",
  adapter    = adapter,
  create     = TRUE
)
conn
```

Nested uploads via `gdpins_raw_put_object()` and `gdpins_raw_put_file()` create
missing intermediate Drive folders automatically (for example
`"sub/sub/folder/file.rds"`).

## Removal behavior (full vs partial)

`gdpins` supports explicit, idempotent remove verbs:

```{r, eval=FALSE}
# Boards: removes a pin from all configured board layers
gdpins_pin_remove(board, "cars")

# Raw connections: removes one file (flat or nested path)
gdpins_raw_remove(conn, "sub/sub/file.rds")
```

Drive removals are recoverable: files are moved to Drive trash (not hard-deleted).
For raw connections, removal scope is single files only; recursive folder deletion
is intentionally unsupported.

## Testing with a fake adapter

In scripts and packages, use `gdpins_fake_drive()` for offline testing:

```{r}
adapter <- gdpins_fake_drive()

board <- gdpins_init_board(
  name       = "test_board",
  drive_path = "project/data-raw",
  cache_dir  = tempfile("cache_"),
  adapter    = adapter,
  create     = TRUE
)
board
```

The fake adapter mirrors Drive operations on the local filesystem — no credentials, no network, fully reproducible.

Note that `gdpins_fake_drive()` is a **test seam**, not an offline-mode substitute: it starts from an empty tempdir, so swapping it in for `gdpins_real_drive()` on a board or connection you have already been working with will not preserve the existing Drive-side layout, and typically surfaces as create-confirm errors. To temporarily disconnect a real board or connection from Drive, see "Deliberately going offline and back online" below instead.

## Deliberately going offline and back online

`gdpins_init_board()` and `gdpins_raw_connect()` fall back to local-only automatically when Drive is unreachable (see the "Offline behaviour" section of `vignette("gdpins-usage")`). Sometimes you want the same effect on purpose — for example, Drive is flaky, you are working on a plane, or you just don't want network calls slowing down every read/write for a while.

`gdpins_go_offline()` detaches Drive from an already-connected board or raw connection, without touching any files:

```{r offline-mode}
adapter <- gdpins_fake_drive()
board <- gdpins_init_board(
  name       = "data_raw",
  drive_path = "project/data-raw",
  cache_dir  = tempfile("cache_"),
  local_dir  = tempfile("local_"),
  adapter    = adapter,
  create     = TRUE
)

board_offline <- gdpins_go_offline(board)
board_offline$config   # "local_only"

# Reads and writes stay local -- no network calls, no blocked writes
gdpins_pin_write(board_offline, mtcars, "cars")
gdpins_pin_read(board_offline, "cars")
```

Under the hood, `gdpins_go_offline()` reuses whichever local storage the board already had — the standalone `local_dir`/`local_board` for a `"drive_cache_local"` board, or the `cache_dir`/`cache_board` for a `"drive_cache"` board with no standalone local directory (mirroring the automatic offline fallback). Nothing is copied or deleted; the returned object just stops touching Drive. The original Drive configuration travels with the returned object, so `gdpins_go_online()` can restore it later:

```{r online-mode, eval=FALSE}
# Later, once Drive is reachable again:
board_online <- gdpins_go_online(
  board_offline,
  on_discrepancy = "sync_to_drive"   # push what changed while offline
)
board_online$config   # back to "drive_cache_local"
```

`gdpins_go_online()` requires `gdpins_is_online()` to be `TRUE` and errors if `x` was never produced by `gdpins_go_offline()` — it restores a *specific* prior connection rather than guessing at one. `on_discrepancy` takes the same values as `gdpins_init_board()`/`gdpins_raw_connect()` (`"prompt"`, `"warn"`, `"sync_from_drive"`, `"sync_to_drive"`, `"ignore"`), so you control exactly how offline changes get reconciled with Drive. If your Drive credentials expired while offline, pass a freshly authenticated adapter: `gdpins_go_online(board_offline, adapter = gdpins_real_drive("folder-id"))`.

The same two functions work on raw connections (`gdpins_raw_connect()`) the same way — `gdpins_go_offline(conn)` / `gdpins_go_online(conn)`.
