---
title: "Working with Raw Files: Writing, Reading, Listing, and Local Downloads"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Raw Files: Writing, Reading, Listing, and Local Downloads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # Drive operations need auth; keep chunks non-executable in CI
)
```

The raw-exogenous layer in `gdpins` manages files as received from external
sources — APIs, vendor exports, survey downloads — without a pins metadata
wrapper. This vignette covers the full lifecycle: **writing**, **reading**,
**listing**, and **downloading** raw files, including files with non-standard
names.

## Setup

```{r load}
library(gdpins)
```

```{r adapter}
# Authenticate once (picks up GDRIVE_EMAIL from .Renviron)
gdpins_ensure_drive_auth()

# Point the adapter at your project root folder on Drive
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")
```

```{r connect}
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",          # relative to adapter root
  local_path = "data/raw/worldbank-api", # local mirror directory
  adapter    = adapter,
  create     = TRUE
)
conn
```

---

## Writing raw files

There are two ways to store files: serialise an R object, or upload a file
verbatim.

### Write an R object — `gdpins_raw_put_object()`

The extension in `name` controls the serialisation format. Supported formats:

| Extension | Writer | Reader | Notes |
|-----------|--------|--------|-------|
| `.parquet` | `arrow::write_parquet()` | `arrow::read_parquet()` | Columnar, fast, compact. Preferred for tibbles. |
| `.csv` | `readr::write_csv()` | `readr::read_csv()` | Plain text, universally readable. |
| `.rds` | `saveRDS()` | `readRDS()` | Any R object, R-only format. |
| `.geojson` | `sf::st_write()` | `sf::st_read()` | Spatial vector data. |

```{r put-object}
library(dplyr)

# Tibble → Parquet (fast, columnar)
gdpins_raw_put_object(conn, gdp_panel_tbl, "gdp_2024.parquet")

# Tibble → CSV (plain text)
gdpins_raw_put_object(conn, metadata_tbl, "metadata.csv")

# Any R object (list, model, etc.) → RDS
gdpins_raw_put_object(conn, my_model, "model.rds")

# sf object → GeoJSON
gdpins_raw_put_object(conn, parcels_sf, "parcels.geojson")

# sf object → Parquet (geometry encoded as WKT, CRS preserved)
gdpins_raw_put_object(conn, parcels_sf, "parcels.parquet")

# Nested paths are created automatically
gdpins_raw_put_object(conn, api_response, "worldbank/gdp_by_country_2023.parquet")
```

Files are written to both the **local mirror** and **Drive** in one step.

### Upload a file verbatim — `gdpins_raw_put_file()`

Use this when you already have a file on disk and want to upload it byte-faithful,
without any R round-trip (e.g. a downloaded CSV, a shapefile ZIP, a JSON blob):

```{r put-file}
# Upload a file exactly as-is
gdpins_raw_put_file(conn,
  path = "downloads/kadaster_raw.geojson",
  name = "kadaster_raw.geojson"
)

# Non-standard name — no escaping needed
gdpins_raw_put_file(conn,
  path = "downloads/quarterly report (Q1 2024).csv",
  name = "quarterly report (Q1 2024).csv"
)

# Nested destination
gdpins_raw_put_file(conn,
  path = "exports/final/merged.parquet",
  name = "exports/final/merged.parquet"
)
```

---

## Reading raw files

### Read and deserialise — `gdpins_raw_get()`

`gdpins_raw_get()` reads from the **local mirror** and deserialises using the
format matched by the file extension. Use `force_refresh = TRUE` to pull a
fresh copy from Drive first.

```{r raw-get}
# Read from local mirror (default — no network needed if already synced)
gdp_tbl   <- gdpins_raw_get(conn, "gdp_2024.parquet")     # → tibble via arrow
meta_tbl  <- gdpins_raw_get(conn, "metadata.csv")          # → tibble via readr
my_model  <- gdpins_raw_get(conn, "model.rds")             # → R object via readRDS
parcels   <- gdpins_raw_get(conn, "parcels.geojson")       # → sf via sf::st_read

# sf stored as parquet: geometry and CRS are automatically restored
parcels   <- gdpins_raw_get(conn, "parcels.parquet")       # → sf

# Force a re-download from Drive before reading
gdp_fresh <- gdpins_raw_get(conn, "gdp_2024.parquet", force_refresh = TRUE)

# Non-standard filenames work identically
q1_tbl <- gdpins_raw_get(conn, "quarterly report (Q1 2024).csv")
```

### Read via local path

Sometimes you need the **file path** rather than the parsed R object — to pass
to a specialised reader, an external tool, or when the format is not one of the
four built-in ones. Use `gdpins_raw_path()`:

```{r raw-path-read}
# Get path (downloads from Drive if not yet local)
path <- gdpins_raw_path(conn, "gdp_2024.parquet")

# Use any reader directly
arrow::read_parquet(path)
data.table::fread(path)
vroom::vroom(path)

# Non-standard formats — gdpins_raw_get() doesn't parse these, use the path
path_xlsx <- gdpins_raw_path(conn, "source data (raw).xlsx")
readxl::read_excel(path_xlsx)
```

---

## Listing files with `gdpins_raw_ls()`

`gdpins_raw_ls()` returns an 8-column tibble describing everything inside the
raw-root up to `depth` directory levels.

```{r raw-ls}
listing <- gdpins_raw_ls(conn, depth = 2)
listing
```

The eight columns are:

| Column | Type | Description |
|--------|------|-------------|
| `name` | chr | Relative path within the raw-root |
| `is_dir` | lgl | `TRUE` for directories |
| `size` | dbl | File size in bytes |
| `mtime` | POSIXct | Last-modified time |
| `depth` | int | Directory depth (1 = top-level) |
| `local_path` | chr | **Absolute path on this machine** |
| `drive_id` | chr | Google Drive file/folder ID |
| `drive_url` | chr | Clickable browser URL |

### Use the listing to read files directly

```{r local-path-use}
# Filter to a specific file and get its local path
row <- listing[listing$name == "gdp_2024.parquet", ]
local_path <- row$local_path

# Pass the local path straight to any reader
arrow::read_parquet(local_path)
readr::read_csv(local_path)
sf::st_read(local_path)
```

### Open files on Google Drive

```{r drive-url-use}
# Inspect Drive IDs and URLs for every listed file
listing[!listing$is_dir, c("name", "drive_id", "drive_url")]

# Open a file directly in the browser (copy the URL)
listing$drive_url[listing$name == "gdp_2024.parquet"]
```

---

## Downloading a file with `gdpins_raw_path()`

`gdpins_raw_path(conn, name_or_id)` returns the **absolute local path** to a
file, downloading it from Drive first if it is not already present locally.
Files already in the local mirror are returned immediately — no re-download.

```{r raw-path-basic}
# File already in local mirror → instant return
path <- gdpins_raw_path(conn, "gdp_2024.parquet")
file.exists(path)  # TRUE

# File not yet local → downloaded automatically, then path returned
path2 <- gdpins_raw_path(conn, "api/gdp_2023.parquet")
arrow::read_parquet(path2)
```

### Non-standard filenames

Files with spaces, parentheses, hyphens, and other special characters work
without any escaping:

```{r raw-path-nonstandard}
# Spaces and parentheses in filename
path <- gdpins_raw_path(conn, "quarterly report (Q1 2024).csv")
readr::read_csv(path)

# Hyphens and mixed case
path <- gdpins_raw_path(conn, "World Bank - GDP Panel 2000-2023.parquet")
arrow::read_parquet(path)

# Deeply nested path with spaces
path <- gdpins_raw_path(conn, "surveys/round 3/respondents - final.rds")
readRDS(path)
```

### Fetch by Google Drive file ID

If you have a Drive file ID (obtained from `gdpins_raw_ls()$drive_id` or from
the Drive URL), you can fetch the file directly by ID. This is useful when you
know the ID but not the exact relative path.

```{r raw-path-id}
# Get IDs from the listing
listing   <- gdpins_raw_ls(conn)
file_id   <- listing$drive_id[listing$name == "gdp_2024.parquet"]

# Fetch by ID — downloads to conn$local_path using original filename
local_path <- gdpins_raw_path(conn, file_id)
arrow::read_parquet(local_path)
```

> **Note:** Drive ID lookup requires a real adapter (`gdpins_real_drive()`).
> It is not supported on local-only connections or the fake adapter used in
> tests.

---

## Workflow: listing → path → read

A complete pattern for reading all files of a given type from Drive:

```{r workflow-all}
listing <- gdpins_raw_ls(conn, depth = 3)

# Filter to parquet files only
parquet_files <- listing[!listing$is_dir & grepl("\\.parquet$", listing$name), ]

# Download each (skips files already local) and read
datasets <- lapply(parquet_files$name, function(name) {
  path <- gdpins_raw_path(conn, name)
  arrow::read_parquet(path)
})
names(datasets) <- parquet_files$name
```

## Workflow: force-refresh the local mirror

To pull all Drive files to the local mirror in one shot (e.g. on a new
machine or after a Drive update):

```{r force-refresh}
gdpins_refresh_disconnect(conn)
# All Drive files are now present in conn$local_path
# Re-connect afterwards if you need to keep using the connection
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = "data/raw/worldbank-api",
  adapter    = adapter
)
```

---

## Testing without a network connection

All the above works identically against the fake adapter used in tests and
offline development — the only difference is that `drive_id` and `drive_url`
will be `NA_character_` (the fake filesystem has no real Drive IDs).

```{r fake-adapter, eval = TRUE}
library(gdpins)

adapter <- gdpins_fake_drive()
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = tempfile("raw_"),
  adapter    = adapter,
  create     = TRUE
)

# Write files in various formats
gdpins_raw_put_object(conn, mtcars,  "cars.parquet")
gdpins_raw_put_object(conn, mtcars,  "cars.csv")
gdpins_raw_put_object(conn, mtcars,  "cars.rds")
gdpins_raw_put_object(conn, mtcars,  "quarterly report (Q1 2024).csv")

# Read back by format
gdpins_raw_get(conn, "cars.parquet")  # tibble via arrow
gdpins_raw_get(conn, "cars.csv")      # tibble via readr
gdpins_raw_get(conn, "cars.rds")      # data.frame via readRDS

# Read non-standard filename
gdpins_raw_get(conn, "quarterly report (Q1 2024).csv")

# List — 8-column tibble
tbl <- gdpins_raw_ls(conn)
tbl[, c("name", "local_path", "drive_id")]

# Resolve to local path (already cached — instant)
path <- gdpins_raw_path(conn, "cars.parquet")
arrow::read_parquet(path) |> head(3)

# Remove a file
gdpins_raw_remove(conn, "cars.csv")
```


## Setup

```{r load}
library(gdpins)
```

```{r adapter}
# Authenticate once (picks up GDRIVE_EMAIL from .Renviron)
gdpins_ensure_drive_auth()

# Point the adapter at your project root folder on Drive
adapter <- gdpins_real_drive("1BxiMVs0XRA5nFMdKvBdBZjgmUUqptlbs74OgVE2upms")
```

```{r connect}
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",          # relative to adapter root
  local_path = "data/raw/worldbank-api", # local mirror directory
  adapter    = adapter,
  create     = TRUE
)
conn
```

---

## Listing files with `gdpins_raw_ls()`

`gdpins_raw_ls()` returns an 8-column tibble describing everything inside the
raw-root up to `depth` directory levels.

```{r raw-ls}
listing <- gdpins_raw_ls(conn, depth = 2)
listing
```

The eight columns are:

| Column | Type | Description |
|--------|------|-------------|
| `name` | chr | Relative path within the raw-root |
| `is_dir` | lgl | `TRUE` for directories |
| `size` | dbl | File size in bytes |
| `mtime` | POSIXct | Last-modified time |
| `depth` | int | Directory depth (1 = top-level) |
| `local_path` | chr | **Absolute path on this machine** |
| `drive_id` | chr | Google Drive file/folder ID |
| `drive_url` | chr | Clickable browser URL |

### Use the listing to read files directly

```{r local-path-use}
# Filter to a specific file and get its local path
row <- listing[listing$name == "gdp_2024.parquet", ]
local_path <- row$local_path

# Pass the local path straight to any reader
arrow::read_parquet(local_path)
readr::read_csv(local_path)
sf::st_read(local_path)
```

### Open files on Google Drive

```{r drive-url-use}
# Inspect Drive IDs and URLs for every listed file
listing[!listing$is_dir, c("name", "drive_id", "drive_url")]

# Open a file directly in the browser (copy the URL)
listing$drive_url[listing$name == "gdp_2024.parquet"]
```

---

## Downloading a file with `gdpins_raw_path()`

`gdpins_raw_path(conn, name_or_id)` returns the **absolute local path** to a
file, downloading it from Drive first if it is not already present locally.
Files already in the local mirror are returned immediately — no re-download.

```{r raw-path-basic}
# File already in local mirror → instant return
path <- gdpins_raw_path(conn, "gdp_2024.parquet")
file.exists(path)  # TRUE

# File not yet local → downloaded automatically, then path returned
path2 <- gdpins_raw_path(conn, "api/gdp_2023.parquet")
arrow::read_parquet(path2)
```

### Non-standard filenames

Files with spaces, parentheses, hyphens, and other special characters work
without any escaping:

```{r raw-path-nonstandard}
# Spaces and parentheses in filename
path <- gdpins_raw_path(conn, "quarterly report (Q1 2024).csv")
readr::read_csv(path)

# Hyphens and mixed case
path <- gdpins_raw_path(conn, "World Bank - GDP Panel 2000-2023.parquet")
arrow::read_parquet(path)

# Deeply nested path with spaces
path <- gdpins_raw_path(conn, "surveys/round 3/respondents - final.rds")
readRDS(path)
```

### Fetch by Google Drive file ID

If you have a Drive file ID (obtained from `gdpins_raw_ls()$drive_id` or from
the Drive URL), you can fetch the file directly by ID. This is useful when you
know the ID but not the exact relative path.

```{r raw-path-id}
# Get IDs from the listing
listing   <- gdpins_raw_ls(conn)
file_id   <- listing$drive_id[listing$name == "gdp_2024.parquet"]

# Fetch by ID — downloads to conn$local_path using original filename
local_path <- gdpins_raw_path(conn, file_id)
arrow::read_parquet(local_path)
```

> **Note:** Drive ID lookup requires a real adapter (`gdpins_real_drive()`).
> It is not supported on local-only connections or the fake adapter used in
> tests.

---

## Workflow: listing → path → read

A complete pattern for reading all files of a given type from Drive:

```{r workflow-all}
listing <- gdpins_raw_ls(conn, depth = 3)

# Filter to parquet files only
parquet_files <- listing[!listing$is_dir & grepl("\\.parquet$", listing$name), ]

# Download each (skips files already local) and read
datasets <- lapply(parquet_files$name, function(name) {
  path <- gdpins_raw_path(conn, name)
  arrow::read_parquet(path)
})
names(datasets) <- parquet_files$name
```

## Workflow: force-refresh the local mirror

To pull all Drive files to the local mirror in one shot (e.g. on a new
machine or after a Drive update):

```{r force-refresh}
gdpins_refresh_disconnect(conn)
# All Drive files are now present in conn$local_path
# Re-connect afterwards if you need to keep using the connection
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = "data/raw/worldbank-api",
  adapter    = adapter
)
```

---

## Testing without a network connection

All the above works identically against the fake adapter used in tests and
offline development — the only difference is that `drive_id` and `drive_url`
will be `NA_character_` (the fake filesystem has no real Drive IDs).

```{r fake-adapter, eval = TRUE}
library(gdpins)

adapter <- gdpins_fake_drive()
conn <- gdpins_raw_connect(
  drive_path = "worldbank-api",
  local_path = tempfile("raw_"),
  adapter    = adapter,
  create     = TRUE
)

# Put some files (including non-standard names)
gdpins_raw_put_object(conn, mtcars, "cars.csv")
gdpins_raw_put_object(conn, mtcars, "quarterly report (Q1 2024).csv")

# List
tbl <- gdpins_raw_ls(conn)
tbl[, c("name", "local_path", "drive_id")]

# Resolve path (already local — instant)
path <- gdpins_raw_path(conn, "cars.csv")
read.csv(path) |> head(3)
```
